---
title: "Analysis of Population Dynamics by Age Groups and Regions"
author: "Bailey Dalton, Dhanya Dharshini Veeraraghavan, Vincent Currie"
date: "2024-11-20"
output: html_document
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(plotly)
```

## Data Preperation

```{r}
# Load the dataset
data <- read.csv("Data/population-young-working-elderly.csv")

```

## Visualisation

#### Fig 1: Population Aged 65+ for Selected countries (1950-2020) - Dhanya

```{r}

# Define countries to highlight
highlight_countries <- c("United States", "China", "India", "Australia", "Japan")

# Filter data for highlighted countries
highlighted_data <- data %>% filter(Entity %in% highlight_countries)

# Create the stacked area chart
ggplot(highlighted_data, aes(x = Year, y = Population_65plus, fill = Entity)) +
  geom_area(alpha = 0.8) +  # Use geom_area for stacked chart
  labs(title = "Population Aged 65+ for Selected countries",
       x = "Year",
       y = "Population",
       fill = "Country") +
  theme_minimal() +
  scale_fill_manual(values = c(
    "United States" = "red", 
    "China" = "yellow", 
    "India" = "green", 
    "Australia" = "purple", 
    "Japan" = "orange"
  ))

```

#### Population Aged 0-14 for Selected Countries (1950-2020)

```{r}

# Define countries to include
selected_countries <- c("United States", "China", "India", "Australia", "Japan")

# Filter data for the selected countries
filtered_data <- data %>%
  filter(Entity %in% selected_countries) %>%
  group_by(Entity, Year) %>%
  summarise(Population_0to14 = sum(Population_0to14, na.rm = TRUE))

# Create the stacked area chart
ggplot(filtered_data, aes(x = Year, y = Population_0to14, fill = Entity)) +
  geom_area(alpha = 0.8) +  # Stacked area chart
  labs(title = "Population Aged 0-14 for Selected Countries (1950-2020)",
       x = "Year",
       y = "Population",
       fill = "Country") +
  theme_minimal() +
  scale_fill_manual(values = c(
    "United States" = "red", 
    "China" = "blue", 
    "India" = "green", 
    "Australia" = "purple", 
    "Japan" = "orange"
  ))

```

## Tables

```{r}
# Calculate growth in the population aged 65+
growth <- data %>%
  group_by(Entity) %>%
  summarise(Start_1950 = first(Population_65plus),
            End_2023 = last(Population_65plus)) %>%
  mutate(Growth = End_2023 - Start_1950) %>%
  arrange(desc(Growth)) %>%
  head(5)

# Display the table
kable(growth, caption = "Table 1: Top 5 Countries with Highest Growth in Population Aged 65+")
```

```{r}
# Calculate growth in the population aged 0-14 for each country
growth <- data %>%
  group_by(Entity) %>%
  summarise(Start_1950 = first(Population_0to14),
            End_2023 = last(Population_0to14),
            Growth = End_2023 - Start_1950) %>%
  arrange(desc(Growth)) %>%
  head(5)

# Display the table
kable(growth, caption = "Table 2: Top 5 Countries with Highest Growth in Population Aged 0-14")
```

## Analysis


###### Figure 1: Population Aged 65+ for Selected Countries (1950-2020)

This chart highlights the rapid growth of elderly populations in selected countries. China shows the steepest increase due to improved healthcare and longer life expectancy (World Bank, 2023). Japan’s sharp rise post-1970 reflects its aging crisis caused by low fertility and high longevity, leading to challenges in sustaining healthcare and pensions (OECD, 2023). India’s slower growth suggests it remains in an earlier demographic stage, while the United States and Australia exhibit steady but less pronounced growth.

###### Figure 2: Population Aged 0-14 for Selected Countries (1950-2020)

India retains the largest 0-14 population due to high fertility rates (UNICEF, 2024). In contrast, China’s decline post-1970s aligns with its one-child policy (Pletcher, 2024). Developed nations like the United States, Japan, and Australia show consistent declines, reflecting lower birth rates and the challenges of maintaining a sustainable workforce in aging populations.

###### Table 1: Top 5 Countries with Highest Growth in Population Aged 65+

China, India, and Japan lead in growth, driven by rising life expectancies and large populations (UN DESA, 2023). Brazil and Russia follow, reflecting similar aging trends in emerging economies as they transition demographically.

###### Table 2: Top 5 Countries with Highest Growth in Population Aged 0-14

India, Nigeria, and Pakistan dominate, driven by high fertility rates in developing nations. Indonesia’s rise signals Southeast Asia’s growing youth population, while China’s moderated growth reflects earlier demographic trends during its high-fertility era.

## Conclusion - Append this to final conclusion:

These trends illustrate aging challenges in developed nations and youth-driven opportunities in developing regions. Targeted policies in healthcare, education, and economic planning are essential to manage these demographic shifts effectively.

## References

World Bank. (2023). Population and demographic indicators. Retrieved from https://www.worldbank.org

OECD. (2023). The silver economy: Policies for an aging society. Retrieved from https://www.oecd.org

UNICEF. (2024). State of the world's children. Retrieved from https://www.unicef.org

Pletcher, K. (2024). China's one-child policy and its effects. Encyclopaedia Britannica. Retrieved from https://www.britannica.com

United Nations Department of Economic and Social Affairs. (2023). World population prospects. Retrieved from https://population.un.org