---
title: "Analysis of Population Dynamics by Age Groups and Regions"
author: "Bailey Dalton, Dhanya Dharshini Veeraraghavan, Vincent Currie"
date: "`r toupper(format(Sys.Date(), '%d-%b-%Y'))`" 
output: 
  bookdown::html_document2:
    css: !expr here::here("Scripts","custom.css") # Examples taken from https://www.w3schools.com/css/default.asp.
    toc: true
    df_print: paged
    number_sections: true
    fig_caption: true
bibliography: references.bib
csl: apa.csl
---

```{r Setup, message=TRUE, warning=TRUE}

#
# Set the default behaviour for displaying code in all chunks to ensure that all display their code, 
# unless turned off within a specific chunk.
#
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r Libs, include=FALSE}
# Load necessary libraries.
#
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(plotly)
# Load the dataset
library(readr)  # The read_csv() command is more efficient than base R read.csv() for loading:
                #   https://bookdown.org/rdpeng/rprogdatascience/using-the-readr-package.html,
                #   https://rforhr.com/index.html.
library(here)   # Ensure that file paths are always relative to the project root directory:
                #   https://cran.r-project.org/web/packages/here/vignettes/rmarkdown.html.
library(tibble) 
library(scales)
library(kableExtra) 
library(bookdown)
library(reshape2)

```

# Introduction 

This report examines the evolution of gender gaps in education, from primary to tertiary levels, through historical and modern contexts, analysing regional progress, policy impacts, and links to socioeconomic development. It highlights global trends and contrasts between developed and developing countries and explores STEM-specific disparities, including Australia’s gender parity across STEM and non-STEM fields.

The authors have undertaken specific parts of the analysis: Dhanya Dharshini explores historical trends, Bailey Dalton investigates regional progress, and Vince Currie examines STEM-specific disparities. These individual sections come together in the conclusion to provide a comprehensive understanding of the gender gap research questions and their broader implications.

## Research Questions  

This section investigates critical research questions about gender gaps in education across various levels and fields. It examines global trends, regional differences, and how policies influence gender parity. The analysis emphasizes disparities in STEM and non-STEM fields, focusing on Australia’s efforts to achieve equity in educational representation.

### Research Questions - Vince Currie 

1. *How does the gender gap in STEM education compare to other fields globally between 2015 and 2022?* (Author: Vince Currie) This analysis draws on data from the @RN21 to examine gender disparities in education, focusing on the number of enrolled students, new entries, and graduates. It assesses female representation in STEM fields relative to other areas, aiming to determine whether STEM fields experience a shortfall of females. 

    Table \@ref(tab:table-FldCmpr) provides a detailed breakdown of various vocational areas. STEM fields are highlighted in <font size="4">*italics*</font> and <font size="4">large font</font>, while the broader categories of **STEM** and **OTHER** represent collections of different vocational groupings. STEM includes the highlighted fields, while OTHER covers all remaining professions. 
    
    The analysis calculates the gender gap as the difference between the number of males and females in each global category (males - females). A positive value indicates a shortfall of females in STEM (males outnumber females), while a negative value reflects more females than males, suggesting no shortfall.

2. *How does gender parity in STEM education compare to non-STEM fields in Australia across different levels and fields from 2015 to 2022?* (Author: Vince Currie) This analysis narrows its focus to Australia, drawing on data from the @RN21 to examine gender disparities in education across different levels (e.g., Bachelor's, Master's, and Vocational) and fields of study. It evaluates gender parity by comparing the representation of females and males in STEM fields to non-STEM fields (OTHER) within the Australian context from 2015 to 2022. The aim is to identify specific areas and education levels where gender disparities are most pronounced and assess progress toward achieving parity. 
    
    Table  \@ref(tab:table-FldCmpr) provides a breakdown of various vocational areas in Australia. STEM fields are highlighted in italics and large font, while the OTHER category includes all non-STEM fields. The analysis calculates the gender gap as the difference between the percentage of males and females in each category (males - females). A positive value indicates male dominance (males outnumber females), while a negative value indicates female dominance (females outnumber males). A value close to zero reflects near parity. Near parity is defined as $\|Gender Gap\|\space <= \space 1.0$, meaning the representation of males and females is approximately equal within this range. 

### Reserach Questions - Dhanya

  * How has the population aged 65+ evolved globally over the past few decades, and what regions show the fastest growth in this age group?  
    - Aging populations are growing faster in developed countries due to higher life expectancy and declining fertility rates.
    - Certain developing countries may show a recent sharp increase, indicating early stages of demographic transition.
    - Regional differences highlight the need for targeted healthcare and social support systems.


•	How does the gender gap in STEM (Science, Technology, Engineering, and Mathematics) education compare to other fields of study?
•	What policies or factors have contributed to significant reductions in educational gender gaps?

Dhanya Dharshini explores the evolution of the 65+ population and trends in the 0-14 age group, highlighting demographic shifts. Bailey Dalton investigates how economic development influences population changes across different regions, analyzing trends from 1950 to 2020 in low, medium, and high-income countries. Vince Currie examines gender disparities in STEM education, comparing them to other fields and identifying policies that have reduced these gaps. The findings provide a comprehensive understanding of these key research questions and their broader implications.

### Research Questions - Bailey 

1. *How has the population distribution across continents changed over the decades, particularly in terms of different age groups, and what are the potential causes and implications of these trends?*

2. *How has a region’s economic development influenced population changes over time? Specifically, what are the population trends from 1950 to 2020 in low, medium, and high economic regions in each age group (youth, working and elderly)?*


## Data Introduction

The data utilised in this report was retrieved from two sections of [Our World in Data](https://ourworldindata.org/). The first data set from _Population of young, working-age and elderly_ containing population recordings over the time period of 1950 to 2023 for various regions around the globe and is available here - [data1](https://ourworldindata.org/grapher/population-young-working-elderly-with-projections.csv?v=1&csvType=full&useColumnShortNames=true). The second data source used is from  _Population and Demographic Change_ and contains information about enrolment ratios for differing educatino levels around the world by gender and is available here - [data2](https://ourworldindata.org/grapher/gender-gap-education-levels.csv?v=1&csvType=full&useColumnShortNames=true) chart found in the topic on [Our World in Data](https://ourworldindata.org/). 


## Data Preparation 

This section describes the steps to preprocess and organise the data, including cleaning, transforming, and categorising variables. It ensures the data-set's readiness for analysis and aligns with the research objectives.

```{r ProcData, message=TRUE, warning=TRUE}
#
# Load the dataset.
#
fp1 <- here("Data", "population-young-working-elderly.csv")

#
# Load the Population Young Working Elderly CSV file.
#
d1 <- read_csv(fp1, show_col_types = FALSE) 

# rename column variables of d1
d1 <- d1 %>% 
   rename(
    Population_65plus = `Population - Sex: all - Age: 65+ - Variant: estimates`,
    Population_15to64 = `Population - Sex: all - Age: 15-64 - Variant: estimates`,
    Population_0to14 = `Population - Sex: all - Age: 0-14 - Variant: estimates`
  )

#
# Set the relative file path (assuming the file is in the 'Data' directory).
# Use here() to specify the relative file path using the root of the R project 
# (where the .Rproj file is located).
#
fp2 <- here("Data", "Gender_education_area2015_2022.csv")

#
# Load the Gender Education csv.
#
d2 <- read_csv(fp2, show_col_types = FALSE)

#
# Vector of the required columns (Author: Vince Currie).
#
sc2 <- c("Reference area", 
         "Education level", 
         "Field of education", 
         "Sex",
         "OBS_VALUE")

#
# The following steps are undertaken (Author: Vince Currie):
#    Rename the column names to something more meaningful.
#    Obtain the mean percentage by gender category for country, education level and field using group by and summarise.
#    Turn the gender category into columns with the calculated means using pivot: giving "Total" and "Female" columns.
#    Calculate the "Male" %  using mutate to create the column as Male = Total - Female. The coalesce replaces NA with 0.
#    
#
 rn2 <- d2[, sc2] %>%
    rename(Cntry = "Reference area",           # Country.
           Edctn_lvl = "Education level",      # Education Level.
           Fld = "Field of education",         # Field.
           Gndr_ctgry = "Sex",                 # Gender Category (Values: Female and Total)
           Obsrvtn_vl = "OBS_VALUE")  %>%      # Percentage. For example, percentage of female participation.
    group_by(Cntry, Edctn_lvl, Fld, Gndr_ctgry)  %>%
    summarise(mObsrvtn_vl = mean(Obsrvtn_vl,na.rm = TRUE),.groups = "drop") %>%
    pivot_wider(names_from = Gndr_ctgry, values_from = mObsrvtn_vl) %>%
    mutate(Male = coalesce(Total,0) - coalesce(Female,0)) 

# 
# Define a vector of STEM-related keywords to search for to classify field as stem or not (Author: Vince Currie).
#
 stem_keywords <- c("Science", "Technology", "Engineering", "Mathematics", "ICT", "Veterinary")

#
# Add a new column 'STEM_Status' categorizing as 'STEM' or 'OTHER' (Author: Vince Currie).
#    Exclude "Total (excluding unknown)" as its not a discipline.
#    Add STEM_Status to the data frame using mutate:
#       * Collapse stem_keywords down to a string of "or" cases using paste and collapse keyword.
#       * Search the "Fld" column looking for a stem keyword in each row using "grepl" command, ignoring case.
#       *   If a stem keyword is found, assign the STEM_Status column the value "STEM", 
#       *   else assign the STEM_Status column the value "OTHER" using the "case_when" command. 
#       *   Make the special case of "Social sciences" "OTHER" as its not part of stem but "science" is in the name.
#    Change the STEM_Status column to categorical using mutate and,
#    Calculate the Gender Gap which is the difference between Male and Female percentages working in that field:
#       If the Gender Gap is positive, it means that Males are dominant in the field,
#       else it means that females are dominant if the field for Gender Gap < 0.
# 
rn2 <- rn2 %>%
  filter(Fld != "Total (excluding unknown)") %>%  
  mutate(STEM_Status = case_when(
    grepl(paste(stem_keywords, collapse = "|"), Fld, ignore.case = TRUE) & 
    !grepl("social science|arts|humanities", Fld, ignore.case = TRUE) ~ "STEM",
    TRUE ~ "OTHER"
  )) %>%
  mutate(STEM_Status = factor(STEM_Status),   
         Gender_Gap = round(coalesce(Male,0) - coalesce(Female,0),2))          

#
# Get the unique field areas being compared as part of Gender Gap (Author: Vince Currie).
#
unique_fields <- unique(rn2$Fld)

#
# Convert the unique fields into a data frame for display as a formatted table (Author: Vince Currie).
#
unique_fields_df <- data.frame(FieldArea = unique_fields)

#
# Rename the column from 'FieldArea' to 'Field Area' (Author: Vince Currie).
#
colnames(unique_fields_df)[colnames(unique_fields_df) == "FieldArea"] <- "Field Area"

#
# Calculate the summary statistics for gender gap by grouping by STEM_Status (Author: Vince Currie).
# Apply multiple summary functions to 'Gender_Gap' using across within summarise.
#   * ".x" represents the value of "Gender_Gap".
#   * ".names = "{.fn}_Gender_Gap" creates the columns as Lower_Gender_Gap, and so on.
#   * The results are rounded to 2 decimal places using round.
#   * The Lower and Upper are capped at -100 and 100 using pmin and pmax. There's only 100% of something.
#
summary_stats <- rn2 %>%
  group_by(STEM_Status) %>%
  summarise(
    across(
      Gender_Gap, 
      list(
        Lower = ~round(pmax(min(.x, na.rm = TRUE), -100), 2),  # Ensure min is capped at -100
        Upper = ~round(pmin(max(.x, na.rm = TRUE), 100), 2),   # Ensure max is capped at 100
        Median = ~round(median(.x, na.rm = TRUE), 2),
        Mean = ~round(mean(.x, na.rm = TRUE), 2)
      ),
      .names = "{.fn}_Gender_Gap"
    )
  )

#
# Calculate Q1 (25%) and Q3 (75%) for Gender_Gap grouped by STEM_Status (Author: Vince Currie).
#
iqr_stats <- rn2 %>%
  group_by(STEM_Status) %>%
  summarise(
    Q1 = round(quantile(Gender_Gap, 0.25, na.rm = TRUE),2),
    Q3 = round(quantile(Gender_Gap, 0.75, na.rm = TRUE),2)
  )

#
# Australia only rows for visualising the Australia data (Author: Vince Currie):
#   * Filter the Australian data.
#   * Select the required columns.
australia_data <- rn2 %>%
  filter(Cntry == "Australia") %>%                                       
  select(Cntry, Edctn_lvl, Fld, STEM_Status, Gender_Gap)                 

#
# Australia data table (Author: Vince Currie).
#   * Rename columns for display in table.
#   * Select the fields.
#   * Sort the data by Field and Education_Level using arrange.
#   * Display Field once per group using lag and mutate.
#
# The table is used to augment the figure by showing those 
# rows where the gender gap is close to parity (|Gender Gap| <= 1).
#
australia_tab <- australia_data %>%
  rename(Field = Fld,
         Level = Edctn_lvl,
         `Gender Gap` = Gender_Gap,
         `Stem Status` = STEM_Status) %>%
  arrange(`Stem Status`, Field, Level) %>%
  mutate(
    Field = ifelse(is.na(lag(Field)) | lag(Field) != Field, Field, "")
  ) %>%
  filter(abs(`Gender Gap`) <= 1) %>%
  select( Field, `Stem Status`, Level, `Gender Gap`) 

```


# Visualisation


The visualisation section presents key insights through graphs and charts, illustrating gender gaps across education levels, fields, and regions. It highlights disparities in STEM and non-STEM fields, global trends, and Australia's context. These visualisations enhance understanding by showcasing patterns, variability, and progress toward achieving gender parity in education.

## Visualisation - Dhanya

```{r Pop65, fig.cap="Population Aged 65+ for Selected countries (1950-2020) - (Author:Dhanya)", fig.topcaption=TRUE}

# Define countries to highlight
highlight_countries <- c("United States", "China", "India", "Australia", "Japan")

# Filter data for highlighted countries
# VC: highlighted_data <- data %>% filter(Entity %in% highlight_countries)
highlighted_data <- d1 %>% filter(Entity %in% highlight_countries)

# Create the stacked area chart
ggplot(highlighted_data, aes(x = Year, y = Population_65plus, fill = Entity)) +
  geom_area(alpha = 0.8) +  # Use geom_area for stacked chart
  labs(x = "Year",
       y = "Population",
       fill = "Country") +
  theme_minimal() +
  scale_fill_manual(values = c(
    "United States" = "red", 
    "China" = "yellow", 
    "India" = "green", 
    "Australia" = "purple", 
    "Japan" = "orange"
  ))

```


```{r Pop014, fig.cap="Population Aged 0-14 for Selected Countries (1950-2020) - (Author:Dhanya)", fig.topcaption=TRUE}

# Define countries to include
selected_countries <- c("United States", "China", "India", "Australia", "Japan")

# Filter data for the selected countries
# VC: filtered_data <- data %>%
filtered_data <- d1 %>%
  filter(Entity %in% selected_countries) %>%
  group_by(Entity, Year) %>%
  summarise(Population_0to14 = sum(Population_0to14, na.rm = TRUE))

# Create the stacked area chart
ggplot(filtered_data, aes(x = Year, y = Population_0to14, fill = Entity)) +
  geom_area(alpha = 0.8) +  # Stacked area chart
  labs(x = "Year",
       y = "Population",
       fill = "Country") +
  theme_minimal() +
  scale_fill_manual(values = c(
    "United States" = "red", 
    "China" = "blue", 
    "India" = "green", 
    "Australia" = "purple", 
    "Japan" = "orange"
  ))

```


## Visualisation - Vince Currie

```{r Figure-Gngp, warning=FALSE, fig.cap="World STEM Gender Gap(%)<br>(Based on 2015-2022 Data) - (Author: Vince Currie)", fig.topcaption=TRUE}

#
# Visualize the gender gap for STEM and non-STEM Vocations using a box plot with "geom_boxplot".
#
ggplot(rn2, aes(x = STEM_Status, y = Gender_Gap, fill = STEM_Status)) +
  geom_boxplot() +
  
  # Display a circle at the lower and upper ends of the whiskers.
  geom_point(
    data = summary_stats,
    aes(x = STEM_Status, y = Lower_Gender_Gap),
    shape = 21,
    fill = "red",
    size = 4
  ) +
  geom_point(
    data = summary_stats,
    aes(x = STEM_Status, y = Upper_Gender_Gap),
    shape = 21,
    fill = "red",
    size = 4
  ) +
  
  # Display the lower and upper values.
  geom_text(
    data = summary_stats,
    aes(x = STEM_Status, y = Lower_Gender_Gap, label = Lower_Gender_Gap),
    vjust = 3,
    size = 2.5
  ) +
  geom_text(
    data = summary_stats,
    aes(x = STEM_Status, y = Upper_Gender_Gap, label = Upper_Gender_Gap),
    vjust = -1.5,
    size = 2.5
  ) +
  
  # Display the middle 50%.
geom_text(
  data = iqr_stats, 
  aes(x = STEM_Status, y = Q1, label = paste("Q1:", round(Q1, 2))),
  hjust = 0.5,                                # Centre horizontally above the bar.
  vjust = 1.5,                                # Adjust vertically to avoid overlap.
  color = "black", size = 2.5
) +
geom_text(
  data = iqr_stats, 
  aes(x = STEM_Status, y = Q3, label = paste("Q3:", round(Q3, 2))),
  hjust = 0.5,                                # Centre horizontally above the bar.
  vjust = -0.8,                               # Adjust vertically to avoid overlap.
  color = "black", size = 2.5
) +
  
  # Add mean value diamond.
  geom_point(
    data = summary_stats,
    aes(x = STEM_Status, y = Mean_Gender_Gap, shape = "Mean"),
    size = 2,
    color = "black"
  ) +
  
  labs(x = "Field Category", y = "Gender Gap (%) (Male - Female)") +
  
  # A legend for the mean value.
  scale_shape_manual(name = "Statistics", values = c("Mean" = 18)) +
  theme_minimal() +
  
  # Add call out boxes for the mean and median.
  geom_label(data = summary_stats, 
             aes(x = STEM_Status, y = Mean_Gender_Gap, label = paste("Mean: ", round(Mean_Gender_Gap, 2))),
             color = "black", size = 2.5, label.padding = unit(0.5, "lines"), label.size = 0.5, 
             vjust = 4.5, hjust = -0.2, fill = "white") +  # White background for the call out label.
  geom_label(data = summary_stats, 
             aes(x = STEM_Status, y = Median_Gender_Gap, label = paste("Median: ", round(Median_Gender_Gap, 2))),
             color = "black", size = 2.5, label.padding = unit(0.5, "lines"), label.size = 0.5, 
             vjust = 3.4, hjust = -0.2, fill = "white") +  # White background for the call out label.
  
  scale_fill_manual(values = c("STEM" = "lightblue", "OTHER" = "lightcoral")) +
  theme(legend.title = element_blank()) +
  scale_y_continuous(breaks = seq(-110, 40, by = 10)) +  # Adjust y-axis ticks.
  expand_limits(y = c(40,-110)) +                        # Ensure that the plot expands to y = (-110,40). 
  
  theme(legend.title = element_blank(),
    plot.margin = margin(t = 30, r = 10, b = 10, l = 20)
        )

```

```{r Figure-Astrl,warning=FALSE,fig.cap="Gender Gap in STEM Education <br>by<br> Field and Level in Australia (2015–2022) - (Author: Vince Currie)",fig.topcaption=TRUE}

#
# Create a bar chart for Australian STEM education.
# (Useful information on colour: : https://www.r-bloggers.com/2013/09/how-to-expand-color-palette-with-ggplot-and-rcolorbrewer/).
#
ggplot(australia_data, aes(x = Fld, y = Gender_Gap, fill = Edctn_lvl)) +
  geom_bar(stat = "identity", position = "stack", color = "black") +         # Use position = "stack" for stacking.
  scale_fill_brewer(palette = "Set1") +                                      # Distinct colours for education levels.
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) + # Add parity line.
  scale_y_continuous(
    limits = c(-10, 10), 
    breaks = seq(-10, 10, by = 2)                                            # Add more y-axis ticks.
  ) +
  geom_text(
   aes(
    label = ifelse(abs(Gender_Gap) > 1.0, round(Gender_Gap, 2), "")          # Display label only if Gender_Gap > 1.0.
  ),
  position = position_stack(vjust = 0.5),                                    # Centre labels within segments.
  size = 1.8,
  color = "black"
)+
  annotate(
    "text", 
    x = 3, y = 9,                                                            # Adjust the x and y positions.
    label = "(Note: Only values with |Gender Gap| > 1.0 are labeled.)",
    size = 1.5, 
    hjust = 0,                                                               # Align text horizontally.
    color = "black"
  )+
  labs(
    x = "Field of Education",
    y = "Gender Gap (Male - Female) (%)",
    fill = "Education Level"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = -45, 
      hjust = 0, 
      vjust = 0
    ),
    plot.margin = margin(t = 30, r = 10, b = 10, l = 20)
  )

```

## Visualisation - Bailey Dalton 

```{r ContinentPlot, echo = TRUE, eval = TRUE, fig.align='center', fig.cap="Population growth across continents from 1950 to 2020 - (Author: Bailey Dalton)", out.width= "80%", message = FALSE, warning = FALSE, fig.topcaption=TRUE}
Continent_data <- d1 %>% 
  filter(Entity %in% c("Africa (UN)", 
                       "Americas (UN)", 
                       "Asia (UN)", 
                       "Europe (UN)", 
                       "Northern America (UN)", 
                       "Australia")) %>% 
  mutate(Population = Population_65plus + Population_15to64 + Population_0to14)

ggplot(Continent_data, aes(x = Year, y = Population, color = Entity)) +
  geom_col() +
  facet_wrap(~ Entity, scales ="free_y") +
  theme(legend.position = "none") + 
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) + 
  labs(
    title = "Continental population growth from 1950-2023",
    x = "Year", 
    y = "Population",
  ) + 
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.3)
    ) 
```


```{r EcoP, echo = TRUE, eval = TRUE, fig.align='center', fig.cap="Population growth trends across economic regions and age groups - (Author: Bailey Dalton)", out.width= "80%", message = FALSE, warning = FALSE, fig.topcaption=TRUE}
# create data frame for population dev data (socioeco status regions) from data
Population_DevelopmentData <- d1 %>% 
  filter(Entity %in% c("Low-income countries", 
                       "Lower-middle-income countries", 
                       "Upper-middle-income countries", 
                       "High-income countries"
                       )) %>% 
  # rename age group columns for convenience 
  rename(
    "65+" = Population_65plus, 
    "15-64" = Population_15to64, 
    "0-14" = Population_0to14,
    "Socio_Economic_Status" = Entity
  ) %>% 
  # gather data to get it in the correct shape 
  gather(AgeGroup, Population, "65+", "15-64", "0-14") %>% 
  select(!Code) %>%
  mutate(Socio_Economic_Status = factor(Socio_Economic_Status, 
         levels = c("Low-income countries", 
                    "Lower-middle-income countries",
                    "Upper-middle-income countries", 
                    "High-income countries")))
  
# Create socio economic plot with formatting 
ggplot(Population_DevelopmentData, aes(x = Year, y = Population, fill = AgeGroup)) +
  geom_area(alpha = 0.7) + 
  facet_wrap(~ Socio_Economic_Status, nrow = 2) + 
  labs(title = "Population trends from 1950 to 2020 for varying economic regions", 
       x = "Year", 
       y = "Population", 
       fill = "Age Groups") +
  scale_fill_manual(values = c(
    "0-14" = "gold",
    "15-64" = "cyan",
    "65+" = "magenta"
  ))
```

# Tables

The Tables section presents detailed data summaries supporting the report's analysis. It includes gender gap statistics across education levels, fields, and regions, highlighting disparities and progress. These tables provide precise numerical insights, complementing the visualizations and enabling a more profound exploration of global trends, STEM and non-STEM comparisons, and Australia's educational gender dynamics.

## Tables - Dhanya

```{r table-HGrowth, warning=FALSE, tab.cap="Top 5 Countries with Highest Growth in Population Aged 65+ (Author:Dhanya).", tab.cap.style = "inline"}
# Calculate growth in the population aged 65+
growth <- d1 %>%
# growth <- data %>%
  group_by(Entity) %>%
  summarise(Start_1950 = first(Population_65plus),
            End_2023 = last(Population_65plus)) %>%
  mutate(Growth = End_2023 - Start_1950) %>%
  arrange(desc(Growth)) %>%
  head(5)

# Display the table
kable(growth,
      booktabs = TRUE, 
      align = "c", 
      format = "html") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  )
```

```{r table-HGrowth014, warning=FALSE, tab.cap="Top 5 Countries with Highest Growth in Population Aged 0-14 - (Author:Dhanya).", tab.cap.style = "inline"}
# Calculate growth in the population aged 0-14 for each country
growth <- d1 %>%
#growth <- data %>%
  group_by(Entity) %>%
  summarise(Start_1950 = first(Population_0to14),
            End_2023 = last(Population_0to14),
            Growth = End_2023 - Start_1950) %>%
  arrange(desc(Growth)) %>%
  head(5)

# Display the table
kable(growth,
      booktabs = TRUE, 
      align = "c", 
      format = "html") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "center",
    full_width = FALSE
  )
```


## Tables - Vince Currie

```{r  table-FldCmpr, warning=FALSE, tab.cap="Fields Compared for Gender Gap(%) Analysis<br>(Based on 2015-2022 Data) - (Author: Vince Currie).", tab.cap.style = "inline"}

#
# Create the formatted table to show the list of education fields.
# kable options:
# Here is an explanation of the bootstrap options:
#    striped: Adds alternating row colours, which improves readability, especially in long tables.
#    hover: The rows will highlight when you hover over them, making it easier to track position in the table.
#    condensed: The rows are more compact, reducing the vertical space between rows.
# Other kable options:
#    booktabs = TRUE, creates a cleaner, more professional table by replacing the default horizontal lines with well-spaced rules.
#    align = "l", controls the alignment of the columns in the table. The value "l" stands for left alignment. Values can be lcr for left/centre/right.
#    format = "html", the table will be rendered as an HTML table, which can be displayed in a browser.
# kableExtra options:
#    position = "centre", the table will be horizontally centre-aligned within the available space.
#    full_width = FALSE ensures the table doesn't stretch across the page, keeping it compact.
# row_spec, used to align row 0 to the centre(align = "c"), which is the column label for the table. The column name ("Field Area") was left aligned by
# the previous kable step used to left-align (align="l") the Fld text, which was globally applied. Also, the rows 1,5, 8, 9, 10 are italic and large 
# font as these represent STEM areas.
#
knitr::kable(unique_fields_df,
             booktabs = TRUE,
             align = "l",
             format = "html")  %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center",
    full_width = FALSE
  ) %>%
  row_spec(row=0,align = "c") %>% 
  row_spec(1, italic = TRUE, font_size = "large") %>%
  row_spec(5, italic = TRUE, font_size = "large") %>%
  row_spec(8, italic = TRUE, font_size = "large") %>%
  row_spec(9, italic = TRUE, font_size = "large") %>%
  row_spec(10,italic = TRUE, font_size = "large")

```


```{r table-SmmrySttstcs, warning=FALSE, tab.cap="World STEM Gender Gap(%) Statistics<br>(Based on 2015-2022 Data)  - (Author: Vince Currie).", tab.cap.style = "inline"}

#
# Create the formatted table to show the summary statistics.
#
knitr::kable(summary_stats,
             booktabs = TRUE,
             align = "c",
             format = "html")  %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center",
    full_width = FALSE
  )

```


```{r table-Astrltb, warning=FALSE, tab.cap="Gender Gap in STEM Education <br>by<br> Field and Level in Australia (2015–2022) that Approximates Parity  - (Author: Vince Currie)", tab.cap.style = "inline"}

knitr::kable(australia_tab,
             booktabs = TRUE,
             align = "l",
             format = "html")  %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center",
    full_width = FALSE
  )%>%
  row_spec(row=0,align = "c") 

```

## Tables - Bailey Dalton

```{r Continent-summary, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, tab.cap= "Continental population statistics, including populations in 1950 and 2020, growth rates, and age group percentages at both years - (Author: Bailey Dalton)", tab.cap.style = "inline"}
# Create table from continent data 
continent_TabData <- Continent_data %>%
  # filter for years at the start of each decade 
  filter(Year %in% c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020)) %>% 
  # group by country
  group_by(Entity) %>% 
  summarise(
    "Population 1950" = Population[Year == 1950],
    "Population 2020" = Population[Year == 2020],
    "Annual Growth Rate" = ((Population[Year == 2020] / Population[Year == 1950])^(1/70)-1)*100,
    "Percentage Population Aged 0-14 in 1950" = Population_0to14[Year == 1950] / Population[Year == 1950] * 100,
    "Percentage Population Aged 0-14 in 2020" = Population_0to14[Year == 2020] / Population[Year == 2020] * 100,
    "Percentage Population Aged 15-64 in 1950" = Population_15to64[Year == 1950] / Population[Year == 1950] * 100,
    "Percentage Population Aged 15-64 in 2020" = Population_15to64[Year == 2020] / Population[Year == 2020] * 100,
    "Percentage Population Aged 65+ in 1950" = Population_65plus[Year == 1950] / Population[Year == 1950] * 100,
    "Percentage Population Aged 65+ in 2020" = Population_65plus[Year == 2020] / Population[Year == 2020] * 100
            ) %>% 
  # format table values 
  mutate(
    # format population to million
    "Population 1950" = label_number(scale = 1e-6, suffix = "M", accuracy = 0.1)(`Population 1950`), 
    "Population 2020" = label_number(scale = 1e-6, suffix = "M", accuracy = 0.1)(`Population 2020`),
    # Format percentages to 2 decimal places with % symbol 
    "Annual Growth Rate" = paste0(round(`Annual Growth Rate`, 2), "%"),
    "Percentage Population Aged 0-14 in 1950" = paste0(round(`Percentage Population Aged 0-14 in 1950`, 2), "%"),
    "Percentage Population Aged 0-14 in 2020" = paste0(round(`Percentage Population Aged 0-14 in 2020`, 2), "%"), 
    "Percentage Population Aged 15-64 in 1950" = paste0(round(`Percentage Population Aged 15-64 in 1950`, 2), "%"),
    "Percentage Population Aged 15-64 in 2020" = paste0(round(`Percentage Population Aged 15-64 in 2020`, 2), "%"),
    "Percentage Population Aged 65+ in 1950" = paste0(round(`Percentage Population Aged 65+ in 1950`, 2), "%"),
    "Percentage Population Aged 65+ in 2020" = paste0(round(`Percentage Population Aged 65+ in 2020`, 2), "%")
  )

# Create and format summary table for continent data 
knitr::kable(continent_TabData
      ) %>% 
  column_spec(1, bold = TRUE) %>%   # Bold first column 
  column_spec(1, width = "2in") %>%  # increase width of first column for spacing
  column_spec(2, width = "1in") %>%   # adjust width of columsn for values in table 
  kable_styling(bootstrap_options = c("striped", "hover"))   # format table for easy reading 
```


```{r SocioEco, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, tab.cap= "Population trends by age group and income region (1950, 1985, 2020) - (Author: Bailey Dalton)", tab.cap.style = "inline"}
# Rectrieve socioEco from figure 2, data cleanup for table 
Economic_Popdata1 <- Population_DevelopmentData %>% 
  filter(Year %in% c(1950, 1985, 2020)) %>% 
  group_by(Year, Socio_Economic_Status, AgeGroup) %>% 
  summarise(Total_Population = sum(Population, na.rm = TRUE)) %>% 
  ungroup()

# Make the numerical population numbers wide 
Economic_Popdata_wide1 <- Economic_Popdata1 %>%
  spread(key = AgeGroup, value = Total_Population)%>% 
  # Add columns in Eco data for total population and percentage in each age group
  mutate(Total_Population = `0-14` + `15-64` + `65+`,
         "%_0-14" = `0-14` / Total_Population * 100,
         "%_15-64" = `15-64` / Total_Population * 100,
         "%_65+" = `65+` / Total_Population * 100
         ) 

# Select only the numerical population size data for each socioEco region
Economic_populations <- Economic_Popdata_wide1%>%
  select(
    Year, 
    Socio_Economic_Status, 
    `0-14`,
    `15-64`,
    `65+`
  ) %>% 
  # convert format of population for better readability in table 
  mutate(across(starts_with(c("0-14", "15-64", "65+")),
                ~ label_number(scale = 1e-6, suffix = "M", accuracy = 0.1)(.)
                ),
         Year = as.character(Year)
         ) 
# Select the percentage populations and perfoem same actions as above
Economic_pop_percentages <- Economic_Popdata_wide1 %>% 
  select(
    Year, Socio_Economic_Status, `%_0-14`, `%_15-64`, `%_65+`
  ) %>% 
  rename(
    `0-14` = `%_0-14`, `15-64`= `%_15-64`, `65+` = `%_65+`
  ) %>% 
  # Round the percentage to 2 decimal places
  mutate(
    `0-14` = round(`0-14`, 2),  
    `15-64` = round(`15-64`, 2), 
    `65+` = round(`65+`, 2)      
  ) %>% 
  # Convert to string and add "%" sign
  mutate(
    `0-14` = paste0(`0-14`, "%"),  
    `15-64` = paste0(`15-64`, "%"), 
    `65+` = paste0(`65+`, "%")      
  ) %>% 
  # Add % sign to Year column to differentiate percentages data from population num data 
  mutate(Year = paste0(Year, " (%)"))

Economic_Popdata_wide1 <- Economic_populations %>%
  pivot_wider(names_from = Socio_Economic_Status, values_from = c("0-14", "15-64", "65+")) %>% 
  select(
    Year, 
    "0-14_Low-income countries",
    "15-64_Low-income countries", 
    "65+_Low-income countries",
    "0-14_Lower-middle-income countries",
    "15-64_Lower-middle-income countries",
    "65+_Lower-middle-income countries",
    "0-14_Upper-middle-income countries", 
    "15-64_Upper-middle-income countries", 
    "65+_Upper-middle-income countries",
    "0-14_High-income countries",
    "15-64_High-income countries", 
    "65+_High-income countries",
  )

Eco_percentages_wide <- Economic_pop_percentages %>% 
  pivot_wider(names_from = Socio_Economic_Status, values_from = c("0-14", "15-64", "65+"))
  

# Join the two wider tables 
Economic_pop_data_combined <- bind_rows(Economic_Popdata_wide1, Eco_percentages_wide) %>% 
  arrange(Year)


Economic_pop_data_combined %>%
  kable(col.names = c("Year", 
                      "0-14", "15-64", "65+",
                      "0-14", "15-64", "65+",
                      "0-14", "15-64", "65+",
                      "0-14", "15-64", "65+"),
        caption = "Table 2: Population trends by age group and income region (1950, 1985, 2020).") %>%
  kable_styling(full_width = TRUE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%  # Bold the Year column
  row_spec(seq(2, nrow(Economic_pop_data_combined), by = 2), background = "lightgray") %>% 
  add_header_above(c(" " = 1, 
                     "Low-income Countries" = 3, 
                     "Lower-middle-income Countries" = 3, 
                     "Upper-middle-income Countries" = 3, 
                     "High-income Countries" = 3))  
```


# Analysis

## Analysis - Vince Currie

1. *Analysis - How does the gender gap in STEM education compare to other fields globally between 2015 and 2022?* (Author:Vince Currie)

    Figure \@ref(fig:Figure-Gngp) presents a box plot and statistical markers illustrating differences in gender gaps between STEM and OTHER fields.
In STEM fields, the median gender gap is close to zero (`r as.numeric(summary_stats[summary_stats$STEM_Status == "STEM", "Median_Gender_Gap"])`), and the mean (`r as.numeric(summary_stats[summary_stats$STEM_Status == "STEM", "Mean_Gender_Gap"])`) skews slightly negative, indicating modest improvements in gender balance. The relatively narrow interquartile range (IQR), with Q3 (`r as.numeric(iqr_stats[iqr_stats$STEM_Status == "STEM", "Q3"])`) highlighting areas of male dominance and Q1 (`r as.numeric(iqr_stats[iqr_stats$STEM_Status == "STEM", "Q1"])`) reflecting progress in female representation, suggests reduced variability. However, the positive skewness of the box plot underscores persistent male dominance, particularly in disciplines such as engineering and ICT.

    In contrast, OTHER fields display a more pronounced negative mean (`r as.numeric(summary_stats[summary_stats$STEM_Status == "OTHER", "Mean_Gender_Gap"])`) and median (`r as.numeric(summary_stats[summary_stats$STEM_Status == "OTHER", "Median_Gender_Gap"])`), indicating female-majority representation, particularly in health and social sciences. The broader IQR, spanning Q1 (`r as.numeric(iqr_stats[iqr_stats$STEM_Status == "OTHER", "Q1"])`) to Q3 (`r as.numeric(iqr_stats[iqr_stats$STEM_Status == "OTHER", "Q3"])`), suggests more significant variability in gender representation, with a notable skew toward female dominance.

    Additionally, the extremes in Table \@ref(tab:table-SmmrySttstcs) reveal significant male dominance in STEM (`r gsub(" ", "",paste("Upper: ",as.numeric(summary_stats[summary_stats$STEM_Status == "STEM", "Upper_Gender_Gap"]),"%"))`) and instances of female-majority representation (`r gsub(" ", "",paste("Lower: ",as.numeric(summary_stats[summary_stats$STEM_Status == "STEM", "Lower_Gender_Gap"]),"%"))`), with narrower variability than in OTHER fields. OTHER fields show broader extremes (`r gsub(" ", "",paste(as.numeric(summary_stats[summary_stats$STEM_Status == "OTHER", "Upper_Gender_Gap"]),"%"))` to `r gsub(" ", "",paste(as.numeric(summary_stats[summary_stats$STEM_Status == "OTHER", "Lower_Gender_Gap"]),"%"))`), reflecting greater gender diversity.  Figure \@ref(fig:Figure-Gngp) depicts these extremes as red circles, emphasising the range of gender disparities.  These findings underscore the persistent underrepresentation of females in STEM and the critical need for targeted strategies to close the gender gap and foster equity in STEM occupations.

2. Analysis - *How does gender parity in STEM education compare to non-STEM fields in Australia across different levels and fields from 2015 to 2022?* (Author:Vince Currie)

    Figure \@ref(fig:Figure-Astrl) illustrates the gender gap in STEM and non-STEM fields across different education levels in Australia from 2015 to 2022. The stacked bar chart reveals notable disparities, with STEM fields showing consistent male dominance across levels such as Bachelor’s and Master’s, particularly in disciplines like engineering and ICT. However, some fields at the vocational and PhD levels approach gender parity, 
with $∥GenderGap∥\space ≤\space 1.0$ indicating near-equal representation. For example, in the Agriculture, Forestry, Fisheries, and Veterinary category,
all levels except PhD demonstrate approximate parity.  In contrast, females dominate the PhD level (Table \@ref(tab:table-Astrltb)). The table also highlights other fields and education levels where the gender gap is near parity.

    Non-STEM (OTHER) fields demonstrate a contrasting trend, with a pronounced negative gender gap indicating female dominance. For instance, health and education fields exhibit significantly higher female representation across all levels, particularly in Bachelor’s and Master’s degrees (Figure \@ref(fig:Figure-Astrl)). Variability in non-STEM fields reflects broader gender diversity, with some areas achieving parity or female dominance , while others differ significantly. This contrasts with STEM, where male dominance is more uniform and persistent, requiring targeted efforts to address gender disparities. The broader range of gender gap values in non-STEM fields underscores this diversity compared to STEM. These findings emphasize the need for targeted initiatives to encourage female participation in STEM while maintaining equity in non-STEM areas.

## Analysis - Dhanya

### Figure 1: Population Aged 65+ for Selected Countries (1950-2020)

This chart highlights the rapid growth of elderly populations in selected countries. China shows the steepest increase due to improved healthcare and longer life expectancy [@RN16]. Japan’s sharp rise post-1970 reflects its aging crisis caused by low fertility and high longevity, leading to challenges in sustaining healthcare and pensions [@RN17]. India’s slower growth suggests it remains in an earlier demographic stage, while the United States and Australia exhibit steady but less pronounced growth.

### Figure 2: Population Aged 0-14 for Selected Countries (1950-2020)

India retains the largest 0-14 population due to high fertility rates [@RN18]. In contrast, China’s decline post-1970s aligns with its one-child policy [@RN19]. Developed nations like the United States, Japan, and Australia show consistent declines, reflecting lower birth rates and the challenges of maintaining a sustainable workforce in aging populations.

### Table 1: Top 5 Countries with Highest Growth in Population Aged 65+

China, India, and Japan lead in growth, driven by rising life expectancies and large populations (UN DESA, 2023). Brazil and Russia follow, reflecting similar aging trends in emerging economies as they transition demographically.

### Table 2: Top 5 Countries with Highest Growth in Population Aged 0-14

India, Nigeria, and Pakistan dominate, driven by high fertility rates in developing nations. Indonesia’s rise signals Southeast Asia’s growing youth population, while China’s moderated growth reflects earlier demographic trends during its high-fertility era.

## Analysis - Bailey

* **How has population changed through the decades in each continent and what might be the causes for various population trends in the different continents?**

Figure \@ref(fig:ContinentPlot) shows population growth trends have varied significantly across continents from 1950 to 2020. Africa experienced the highest growth rate, with an annual growth rate of `r as.character(continent_TabData[continent_TabData$Entity == "Africa (UN)", "Annual Growth Rate"])`, while Europe saw the lowest at `r as.character(continent_TabData[continent_TabData$Entity == "Europe (UN)", "Annual Growth Rate"])` seen in \@ref(tab:Continent-summary). Africa’s population grew from `r as.character(continent_TabData[continent_TabData$Entity == "Africa (UN)", "Population 1950"])` to `r as.character(continent_TabData[continent_TabData$Entity == "Africa (UN)", "Population 2020"])` (1.38 billion), with a young demographic, as the percentage of those aged 65+ remained low. In contrast, Europe’s elderly population increased significantly, from `r as.character(continent_TabData[continent_TabData$Entity == "Europe (UN)", "Percentage Population Aged 65+ in 1950"])` in 1950 to `r as.character(continent_TabData[continent_TabData$Entity == "Europe (UN)", "Percentage Population Aged 65+ in 2020"])` in 2020, reflecting higher life expectancy and lower birth rates. Asia, while seeing moderate growth in its elderly population (up 5%), maintains a large working-age population. Across all continents, the percentage of those aged 0-14 declined, indicating lower birth rates. These trends are driven by factors like improved healthcare, lower fertility rates, and urbanization. These demographic shifts present challenges for economic growth, healthcare systems, and social services, highlighting the need for targeted policies addressing aging populations and youth employment [@RN22]. 
  
  * **How has a region's economic development influenced population changes over time? Specifically, what are the population trends from 1950 to 2020 in low, medium, and high economic regions in each age group (youth, working and elderly)?"**

The area chart in \@ref(fig:EcoP) and table \@ref(tab:SocioEco) show that low-income regions have experienced minimal growth in the 65+ age group, with its percentage slightly declining from `r as.character(Economic_pop_data_combined[Economic_pop_data_combined$Year == "1950 (%)", "0-14_Low-income countries"])` in 1950 to `r as.character(Economic_pop_data_combined[Economic_pop_data_combined$Year == "2020 (%)", "0-14_Low-income countries"])` in 2020. This likely reflects lower life expectancy and limited healthcare [@RN22]. In contrast, high-income regions saw the elderly population grow from `r as.character(Economic_pop_data_combined[Economic_pop_data_combined$Year == "1950 (%)", "65+_High-income countries"])` to  in 1950 to `r as.character(Economic_pop_data_combined[Economic_pop_data_combined$Year == "2020 (%)", "65+_High-income countries"])` in 2020, indicating better healthcare and longevity. Additionally, both regions saw declines in youth populations, while working-age populations have expanded, especially in high-income areas, highlighting broader demographic shifts driven by economic development.

# Conclusion



In conclusion, the analyses highlight significant trends in gender gaps, demographic shifts, and population dynamics across various regions between 1950 and 2023. While STEM fields globally continue to exhibit male dominance with narrower gender variability compared to non-STEM fields, there are notable differences in gender representation across education levels. In Australia, STEM disciplines show consistent male dominance, but non-STEM fields reveal broader gender diversity and a stronger trend toward female-majority representation. These findings emphasize the need for targeted initiatives to address the gender imbalance in STEM education, fostering greater equity globally and within Australia.

Demographic analysis emphasizes the challenges of an aging global population. Developed nations like China and Japan are experiencing sharp increases in elderly populations due to higher life expectancy, while regions like Africa and Asia see rapid growth driven by high fertility rates. In these areas, the proportion of citizens aged 65 and over remains low, while youth and working-age populations are growing quickly. High-income regions are facing a rise in elderly populations due to improved healthcare and longevity, requiring policies to address this aging trend.

These trends show how demographic shifts, economic development, and gender equity are linked. Low- and middle-income regions see significant growth in youth populations, while high-income regions grapple with aging populations. Targeted policies in healthcare, education, and economic planning are crucial for sustainable growth and societal balance.



# References

